---
title: "p8105_hw2_mk4996"
author: "Miho Kawanami"
date: "2025-09-27"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 999) 
```
# Problem 1

## First: Cleaning the data in pols-month.csv.
```{r}
library(tidyverse)

pols_df = read_csv(file = "./pols-month.csv", 
                          na = c(".", "NA", ""), 
                          show_col_types = FALSE) |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) |>
  mutate(
    month = factor(month.name[month], levels = month.name, ordered = TRUE),
    president = case_when(
      prez_gop == 1 ~ "gop",
      prez_dem == 1 ~ "dem",
      TRUE ~ NA_character_
    )) |>
  select(-day, -prez_dem, -prez_gop)
```

```{r, include=FALSE}
head(pols_df)
```

## Second: Cleaning the data in snp.csv.
```{r}
snp_df = read_csv(file = "./snp.csv",
                         na = c(".", "NA", ""),
                         show_col_types = FALSE) |>
  janitor::clean_names() |>
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) |>
  mutate(
  year = ifelse(year < 20, year + 2000, year + 1900),  
  month = factor(month.name[month], levels = month.name, ordered = TRUE)) |>
  select(year, month, close) |>
  arrange(year, month)
```

```{r, include=FALSE}
  head(snp_df)
```

## Third: Tidying the unemployment data
```{r}
unemp_df = read_csv(file = "./unemployment.csv",
                         na = c(".", "NA", ""),
                         show_col_types = FALSE) |>
  janitor::clean_names() |>
  pivot_longer(-year, names_to = "month", values_to = "unemployment") |>
  mutate(
    month = match(tolower(month), tolower(month.abb)),
    month = factor(month.name[month], levels = month.name, ordered = TRUE)
  ) |>
  arrange(year, month)
```

```{r, include=FALSE}
  head(unemp_df)
```

## Join the datasets
```{r}
pols_snp  = left_join(pols_df, snp_df,  by = c("year","month"))
final_df  = left_join(pols_snp, unemp_df, by = c("year","month")) |>
  arrange(year, month)
```

```{r, include=FALSE}
  head(final_df)
```

## Write a short paragraph about these datasets

The `pols-month` dataset contained the number of national politicians who are democratic or republican at any given time.
The `snp` dataset contained Standard & Poor’s stock market index (S&P), often used as a representative measure of stock market as a whole. 
The `unemployment` dataset reports monthly unemployment percentages.  

After merging, the resulting dataset has `r nrow(final_df)` rows and `r ncol(final_df)` columns, ranging years `r min(final_df$year)` to `r max(final_df$year)`.  
Key variables include `year`, `month`, `president`, `close` (the closing values of the S&P stock index), and `unemployment`.

# Problem2
## Cleaning the data in Mr. Trash Wheel
```{r}
library(readxl)
mr_trash = 
  read_excel("./202409 Trash Wheel Collection Data.xlsx", sheet = "Mr. Trash Wheel", range = "A2:N655") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(year = readr::parse_integer(as.character(year)),
        (sports_balls = as.integer(round(sports_balls))),
        (trash_wheel = "Mr. Trash Wheel")) 
```

```{r, include=FALSE}
head(mr_trash)
```

## Cleaning the data in Professor Trash Wheel
```{r}
prof_trash = 
  read_excel("./202409 Trash Wheel Collection Data.xlsx", sheet = "Professor Trash Wheel", range = "A2:M123") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(trash_wheel = "Professor Trash Wheel")
```

```{r, include=FALSE}
head(prof_trash)
```

## Cleaning the data in Gwynnda Trash Wheel
```{r}
gwynnda = 
  read_excel("./202409 Trash Wheel Collection Data.xlsx", sheet = "Gwynnda Trash Wheel", range = "A2:L266") |> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(trash_wheel = "Gwynnda Trash Wheel")
```

```{r, include=FALSE}
head(gwynnda)
```

## Combine the datasets
```{r}
trash_data = bind_rows(mr_trash, prof_trash, gwynnda)
```

```{r, include=FALSE}
head(trash_data)
```

## Write a short paragraph about these datasets
Mr. Trash Wheel is a water-wheel vessel that intercepts litter and debris in Baltimore’s Inner Harbor; the public spreadsheet records dumpster-level collections, including dates, weights/volumes, and counts of common items. Using those sheets, I combined the Mr., Professor, and Gwynnda data into a single tidy dataset with `r nrow(trash_data)` observations across `r n_distinct(trash_data$trash_wheel)` devices. Each row corresponds to one dumpster and includes fields such as `dumpster`, `date`, `year`, `month`, `weight_tons`, `volume_cubic_yards`, and item counts (e.g., `plastic_bottles`, `polystyrene`, `cigarette_butts`, `glass_bottles`, `plastic_bags`, `wrappers`; `sports_balls` is recorded only on some sheets). For the available data, **Professor Trash Wheel** collected a total of `r trash_data |> filter(trash_wheel == "Professor Trash Wheel") |> summarise(total = sum(weight_tons, na.rm = TRUE)) |> pull(total) |> round(1)` tons of trash, and **Gwynnda** collected `r trash_data |> filter(trash_wheel == "Gwynnda Trash Wheel", year == 2022, str_detect(tolower(month), "^jun")) |> summarise(total = sum(cigarette_butts, na.rm = TRUE)) |> pull(total)` cigarette butts in June 2022.

